The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2014-02-19 paper/ms.tex
Marks New comments

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$
emission line in distant galaxies. We model a galaxy as a sphere with
a homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with maximum
velocities in the range $0-300$ \kms and neutral hydrogen optical
depths in the range $\tau_{\rm H}=10^{5}-10^{7}$. We also consider two
kinds of spatial distribution for the radiation sources with respect
to the sphere: central and homogeneous. Our main finding is that the
line width and the intensity at the line's center increase with
rotational velocity. For homogeneously distributed sources the line
becomes single peaked at rotational velocities larger than the line
width in the static case. Under the same
conditions the escape fraction increases $\sim 30\%$. For radiation
sources located off-center, the line morphology presents a range of
single, double and triple peaks. We show how these results are useful
to interpret recent spectroscopic results of distant $z\sim 2-3$ star
forming galaxies.

@@

The detection of strong \ly emission lines has become an essential
method in extra-galactic astronomy to find distant star-forming
galaxies
\citep{PartridgePeebles,Rhoads00,Gawiser2007,Koehler2007,Ouchi08,Yamada2012,Schenker2012}.
The galaxies detected using this method receive the 
name of \ly emitters (LAEs). A detailed examination of this galaxy
population has diverse implications for galaxy formation, reionization
and the large scale structure of the Universe. Attempts to fully
exploit the physical information included in the \ly line require an
understanding of all the physical factors involved in shaping the
line. Due to the resonant nature of this line, these physical factors
notably include temperature, density and bulk velocity field of the neutral
Hydrogen in the emitting galaxy and its surroundings.

A basic understanding of the quantitative behaviour of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep{Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. Analytic studies of configurations including
some kind of bulk flow only include the case of a sphere with a Hubble
like expansion flow \citep{LoebRybicki}.

A quantitative description of the \ly line has been reached through
Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. In the last
two decades these studies have become popular due to the
availability of computing power. Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}; Later on, the effects of clumpy media
\citep{Hansen06} and expanding/contracting shell/spherical geometries started to
be studied \citep{Verhamme06,Dijkstra06}. Similar codes have applied
these results to semi-analytic models of galaxy formation \citep{Orsi12} and
results of large hydrodynamic simulations
\citep{CLARA,Forero12,Behrens13}. Recently, Monte Carlo codes have also
been applied to the results of high resolution hydrodynamic
simulations of individual
galaxies\citep{Laursen09,Barnes11,Verhamme12,Yajima12}. Meanwhile, recent
developments have been focused on the systematic study of clumpy
outflows \citep{DijkstraKramer}and anisotropic velocity configurations
\citep{Zheng2013}.

@@

Performing that study is the main goal of this paper. We investigate for the
first time the impact of rotation on the morphology of the \ly
line. We focus on a simplified system: a spherical gas cloud with
homogeneous density and solid body rotation, to study the line
morphology and the escape fraction in the presence of dust. We base
our work on two independent Monte Carlo based radiative transfer codes
CLARA \citep{CLARA} and XX \citep{DijkstraKramer} .

@@

For each photon in the simulation we have its initial position inside
the sphere, direction of propagation $\hat{k}_{in}$ and reduced
frequency $x_{in}$. The photon's propagation stops once they cross the
surface of the sphere. At this point we store the position, the outgoing direction
of propagation $\hat{k}_{out}$ and the reduced frequency $x_{out}$. We
define the angle $\Theta$ by $\cos\Theta = \hat{k}\cdot
\hat{k}_{out}$, that is the polar angle of the outgoing photon with
respect to the $z$ axis. Following \cite{Zheng2013} we make the study
of the anisotropic emission in terms of this angle..

@@

\begin{table}
\begin{center}
\begin{tabular}{llc}\hline\hline
Physical Parameter (units) & Symbol & Values\\\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ 50,\ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{a}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\\hline\hline
\end{tabular}
\caption{
  List of the physical parameters that define the spherical models 
  simulated in our Monte Carlo calculations. For each parameter we
  vary the values in the range listed in the last column. Taking into
  account all the possible combinations we end up with $60$ different
  models.} 
\label{table:models}
\end{center}
\end{table}

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central source distribution the average number of
scatterings $\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this set-up the number of scatterings is
proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$, respectively

From Figure \ref{fig:Nscatt} is also clear that for the homogeneous
distribution there is a clear decrease of $\langle N_{\rm
  scatt}\rangle$ as the $V_{\rm max}$ increases. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case.

In this case we find that for a static sphere $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this
represents factors of $(0.66, 0.59, 0.55)$ lower than the centrally
emitted photons. In contrast the analytic solution for the homogeneous
slab gives $0.72$ times less scatterings in the homogeneous case than
in the central one.

@@

In the right panel of Figure \ref{fig:efvsv} we put these results in the context of
the analytic solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{A}$ where $\tau_{A} = (1 - A)\tau_{a}$ , an approximation 
that is valid only in the limit $a\tau_{\rm   H}\gg 1$. The dashed lines in
 the right panel of Figure \ref{fig:efvsv} show the results for the different rotational
velocities for the homogeneous models. First of all we note that the escape fraction does not
increase significantly from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, a fact
that is explained since there is a transition from an opaque to an extremely opaque medium
which affects how the \ly photons escape from the medium.

@@

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; an effect that can be induced by rotation. The same
argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

>>>>

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$
emission line in distant galaxies. We model a galaxy as a sphere with
a homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with maximum
velocities in the range $0-300$ \kms and neutral hydrogen optical
depths in the range $\tau_{\rm H}=10^{5}-10^{7}$. We consider cases 
of a single central \lya source, and a uniform distribution of sources. 
Our main finding are that the escape fraction of \lya photons, and
the intensity at the line center increase with rotational velocity. 
For homogeneously distributed sources the line becomes single peaked 
at rotational velocities larger than the line width in the static case.
For radiation sources located off-center, the line morphology presents 
a range of single, double and triple peaks. We show how these results 
are useful to interpret recent spectroscopic results of distant 
$z\sim 2-3$ star forming galaxies.

@@

The detection of strong \ly emission lines has become an essential
method in extra-galactic astronomy to find distant star-forming
galaxies
\citep{PartridgePeebles,Rhoads00,Gawiser2007,Koehler2007,Ouchi08,Yamada2012,Schenker2012,Finkelstein2013}.
The galaxies detected using this method receive the 
name of \ly emitters (LAEs). A detailed examination of this galaxy
population has diverse implications for galaxy formation, reionization
and the large scale structure of the Universe. Attempts to fully
exploit the physical information included in the \ly line require an
understanding of all the physical factors involved in shaping the
line. Due to the resonant nature of this line, these physical factors
notably include temperature, density and bulk velocity field of the neutral
Hydrogen in the emitting galaxy and its surroundings.

A basic understanding of the quantitative behaviour of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep{Adams72,Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. Analytic studies of configurations including
some kind of bulk flow only include the case of a sphere with a Hubble
like expansion flow \citep{LoebRybicki}.

A quantitative description of the \ly line has been reached through
Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. In the last
two decades these studies have become popular due to the
availability of computing power. Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}; Later on, the effects of clumpy media
\citep{Hansen06} and expanding/contracting shell/spherical geometries started to
be studied \citep{Verhamme06,Dijkstra06,Ahn2014}. Similar codes have applied
these results to semi-analytic models of galaxy formation \citep{Orsi12} and
results of large hydrodynamic simulations
\citep{CLARA,Forero12,Behrens13}. Recently, Monte Carlo codes have also
been applied to the results of high resolution hydrodynamic
simulations of individual
galaxies\citep{Laursen09,Barnes11,Verhamme12,Yajima12}. Meanwhile, recent
developments have been focused on the systematic study of clumpy
outflows \citep{DijkstraKramer}and anisotropic velocity configurations
\citep{Zheng2013}.

@@

Performing that study is the main goal of this paper. We investigate for the
first time the impact of rotation on the morphology of the \ly
line. We focus on a simplified system: a spherical gas cloud with
homogeneous density and solid body rotation, to study the line
morphology and the escape fraction in the presence of dust. We base
our work on two independent Monte Carlo based radiative transfer codes
CLARA \citep{CLARA} and \citep{DijkstraKramer}.

@@

For each photon in the simulation we have its initial position inside
the sphere, direction of propagation $\hat{k}_{in}$ and reduced
frequency $x_{in}$. The photon's propagation stops once they cross the
surface of the sphere. At this point we store the position, the outgoing direction
of propagation $\hat{k}_{out}$ and the reduced frequency $x_{out}$. We
define the angle $\Theta$ by $\cos\Theta = \hat{k}\cdot
\hat{k}_{out}$, that is the polar angle of the outgoing photon with
respect to the $z$ axis. Following \cite{Zheng2013} we make the study
of the anisotropic emission in terms of this angle.

@@

\begin{table}
\begin{center}
\begin{tabular}{llc}\hline\hline
Physical Parameter (units) & Symbol & Values\\\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ 50,\ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{a}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\\hline\hline
\end{tabular}
\caption{
  Summary of Physical Parameters of our Monte Carlo Simulations.} 
\label{table:models}
\end{center}
\end{table}

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central source distribution the average number of
scatterings $\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this set-up the number of scatterings is
proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$, respectively and
$V_{\mathrm{max}}=300$ \kms.

From Figure \ref{fig:Nscatt} for the homogeneous
distribution there is a decrease of $\langle N_{\rm
  scatt}\rangle$ as the $V_{\rm max}$ increases. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case.

For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this
represents factors of $(0.66, 0.59, 0.55)$ lower than the centrally
emitted photons. While for the homogeneous rotating sphere at 
$V_{\rm{max}}=300\kms$ we find that $\langle N_{\rm
  scatt}\rangle=(0.38, 0.34, 0.42)\tau_{\rm   H}$ which represents 
  factors of $(0.26, 0.34, 0.46)$ lower than the corresponding central
  case.  In contrast the analytic solution for the homogeneous
slab gives	 $0.72$ times less scatterings in the homogeneous case than
in the central one

@@

In the right panel of Figure \ref{fig:efvsv} we put these results in the context of
the analytic solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{A}$ where $\tau_{A} = (1 - A)\tau_{a}$ , an approximation 
that is valid only in the limit $a\tau_{\rm   H}\gg 1$. The dashed lines in
 the right panel of Figure \ref{fig:efvsv} show the results for the different rotational
velocities for the homogeneous models. First of all we note that the escape fraction does not
increase significantly from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, This
 counter-intuitive result is a consequence of the transition into the 'extremely' opaque regime,
  which occurs roughly when $a\tau_{\rm{H}} > 10^{3}$ e.g. \citep{Neufeld90}.

@@

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; this could indicate the presence of rotation 
and \lya photons leaking though low column density holes, as in \citep{Behrens2014}.
 The same argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.
